---
title: "cardinalR: Generating interesting high-dimensional data structures"
abstract: >
  A high-dimensional dataset is where each observation is described by many features, or dimensions. Such a dataset might contain various types of structures that have complex geometric properties, such as nonlinear manifolds, clusters, or sparse distributions. We can generate data containing a variety of structures using mathematical functions and statistical distributions. Sampling from a multivariate normal distribution will generate data in an elliptical shape. Using a trigonometric function we can generate a spiral. A torus function can create a donut shape. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package `cardinalR`, along with some already generated examples.

draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: application
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
 rjtools::rjournal_web_article:
    css: "style.css"
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
options(repos = "https://cloud.r-project.org") ## set up CRAN mirror
knitr::opts_chunk$set(
  echo = FALSE,
  cache=FALSE, 
  message=FALSE, 
  warning=FALSE,
  out.width = "100%")

```

```{r load-libraries}
library(tidyverse)
library(knitr)
```

# Introduction

Research gap with `geozoo`, `snedata` packages...

# Implementation

## Installation

The package can be installed from CRAN using 

> ```install.packages("cardinalR")```

and from GitHub using 

> ```remotes::install_github("jayanilakshika/cardinalR")```

to install the development version.

## Web site

More documentation of the package can be found at the web site https://jayanilakshika.github.io/cardinalR/.

## Data sets

The `cardinalR` package comes with several data sets that load with the package. These are described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:datasets-tb-html)', '\\@ref(tab:datasets-tb-pdf)'))`.

```{r}
datasets_tb <- tibble(dt = c("mobius_clust_data", 
                             "mobius_clust_tsne_param1", 
                             "mobius_clust_tsne_param2",
                             "mobius_clust_tsne_param3",
                             "mobius_clust_umap_param1",
                             "mobius_clust_umap_param2",
                             "mobius_clust_umap_param3"), 
                      text = c("",
                               "", 
                               "",
                               "",
                               "",
                               "", 
                               ""))
```

```{r datasets-tb-html, eval=is_html_output()}
datasets_tb |> 
  kable(caption = "cardinalR data sets", col.names = c("data", "explanation")) 
```

```{r datasets-tb-pdf, eval=is_latex_output()}
datasets_tb |> 
  kable(caption = "cardinalR data sets", format="latex", col.names = c("data", "explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

## Functions

### Main function

### Branching

```{r}
branch_tb <- tibble(fun = c("gen_expbranches",
                             "gen_linearbranches",
                             "gen_curvybranches", 
                             "gen_orglinearbranches", 
                             "gen_orgcurvybranches"), 
                      exp = c("",
                               "", 
                               "",
                               "",
                               ""))
```

```{r branching-tb-html, eval=is_html_output()}
branch_tb |> 
  kable(caption = "cardinalR branching data generation functions", col.names = c("Function", "Explanation")) 
```

```{r branching-tb-pdf, eval=is_latex_output()}
branch_tb |> 
  kable(caption = "cardinalR branching data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```


### Cone

### Cube

### Gaussian

### Linear

### Mobius

### Polynomial

### Pyramid

### S-curve

### Sphere

### Swiss Roll  

To generalize the Swiss roll structure to arbitrary dimensions, we introduce a function `generate_swiss_roll(n, p)`, which constructs a high-dimensional version of the classic 3D Swiss roll while preserving its core characteristics.  

The function generates `n` points in a `p`-dimensional space, where the first two dimensions (`X_1, X_2`) define the primary Swiss roll shape using a parametric equation:  

$$
X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad \text{where } t \sim U(0, 3\pi)
$$

The third dimension (`X_3`) introduces variation perpendicular to the roll, sampled uniformly from \([-1,1]\). Additional dimensions (`X_4` to `X_p`) extend the data structure by applying a **sinusoidal transformation** of the parameter `t`, ensuring continuity in higher-dimensional spaces:  

$$
X_i = \frac{\sin(i t)}{i}, \quad \text{for } i \geq 4.
$$

This transformation ensures a gradual decay in variance across dimensions, mimicking real-world high-dimensional structures where later dimensions often capture subtler variations. 

### Odd shapes

### Additional functions

# Examples

Add one or two datasets and evaluate how it will be useful.... (one with NLDR, one with clustering)

# Discussion

# Code

The code is available at <https://github.com/JayaniLakshika/cardinalR>, and source material for this paper is available at <https://github.com/JayaniLakshika/paper-cardinalR>.

# Acknowledgements

This article is created using \CRANpkg{knitr} [@knitr] and \CRANpkg{rmarkdown} [@rmarkdown] in R with the `rjtools::rjournal_article` template. 
